Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-08-10 analysis/paper.qmd
Broaden training region
To West Eurasia, plus related changes suggested by reviewers.

<!--
TODO before publication:
- [ ] Revise discussion of archaeo verification
- [ ] Outstanding TODOs
- [ ] Add missing references
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Zenodo release
- [ ] Final proofread
-->

@@

@@

```{r define-parameters}
# Minimum number archaeological occurrences for modelling
min_n_archaeo <- 1

# Minimum number of present occurrences for modelling
min_n_occ <- 50

@@

Biogeographic research into the centres of origin and/or domestication of crops has also long informed broader understanding of the process of agricultural origins.
In *Man Makes Himself*, @Childe1936 correctly located the centre of origin of European agriculture in the 'Fertile Crescent' of West Asia (unlike for example @Pumpelly1908 before him).
This was not based on the region's prehistoric archaeological record, which at that point was barely known.
Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].
In subsequent decades, the search for more precise origin zones of specific domestic plants relied on the assumption that "the locus of domestication of a wild plant would presumably be within its area of original distribution in the wild state" [@Butzer1971; paraphrasing @Helbaek1959] – and that this "natural habitat" has not changed significantly over the last 12,000 years [@Butzer1971].

@@

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@@

ne_lakes <- read_sf(raw_data("ne"), "ne_10m_lakes") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_union()

ne_ocean <- read_sf(raw_data("ne"), "ne_10m_ocean") |>
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_union()

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_difference(ne_ocean) |>
  # st_difference(ne_lakes) |>
  st_union()

s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)
w_asia_land <- st_intersection(st_as_sfc(w_asia), ne_land)

@@

```{r data-flora}
# Start with flora present at archaeological sites
if (!file_exists(derived_data("flora.Rdata"))) {
  flora <- read_tsv(
    raw_data("swasia_neolithic_flora.tsv"),
    col_types = cols(
      source = col_character(),
      source_id = col_character(),
      source_site_name = col_character(),
      site_name = col_character(),
      latitude = col_double(),
      longitude = col_double(),
      phase = col_character(),
      phase_description = col_character(),
      phase_code = col_character(),
      age_start = col_integer(),
      age_end = col_integer(),
      taxon_source = col_character(),
      n = col_double(),
      prop = col_double(),
      reference = col_character(),
      taxon_detail = col_character(),
      taxon = col_character(),
      genus = col_character(),
      family = col_character(),
      category = col_character(),
      founder_crop = col_character(),
      edibility = col_character(),
      grass_type = col_character(),
      legume_type = col_character()
    )
  ) |>
    rename(taxon_group = "taxon", taxon = "taxon_detail") |>
    # Assign to archaeological periods
    mutate(
      age_mid = age_end + ((age_start - age_end) / 2),
      period = cut(
        -age_mid, 
        breaks = -c(archaeo_periods$start_bp, 0), 
        labels = archaeo_periods$period,
        ordered_result = TRUE
      ),
      climate_period = cut(
        -age_mid, 
        breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
        labels = climate_periods[climate_periods$code != "cur",]$code,
        ordered_result = TRUE
      )
    ) |>
    # Filter to Epipal/Neo
    filter(period %in% epipal | period %in% ppn) |>
    # Group by taxon
    drop_na(taxon) |>
    nest(.by = c(taxon), .key = "archaeo") |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      # Spot corrections
      taxon = case_match(
        taxon,
        c("Bolboschoenus glaucus", "Bolboschoenus maritimus") ~ "Bolboschoenus",
        "Vicia narbonense" ~ "Vicia narbonensis", # N=3
        "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
        "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
        "Lens orientalis" ~ "Vicia orientalis", # N=2
        "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
        "Stipa capensis" ~ "Stipa dregeana", # N=2
        .default = taxon
      ),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name),
      .before = archaeo
    ) |>
    # Drop unmatched (mostly not identified to species)
    drop_na(taxon) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Regroup by canonical taxon
    summarise(
      taxon_aliases = list(taxon_original[taxon_original != taxon]),
      archaeo = list(bind_rows(archaeo)),
      .by = taxon
    ) |>
    mutate(
      n_archaeo = map_int(archaeo, nrow)
    ) |>
    filter(n_archaeo > min_n_archaeo) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    )

@@

```{r fig-region}
#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, fill = NA) +
  layer_spatial(archaeo_sites, size = 0.5) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Based on current understandings, we assume that plant-based subsistence during this period was broad, geographically- and temporally-varied, and reflects a gradual, geographically decentred, and nonlinear tranistion to greater reliance on cultivars (i.e. agriculture, see @sec-bg).
Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@LucasFuller2018; @FullerEtAl2018; based on @ColledgeEtAl2004; @ShennanConolly2007].
We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

The taxonomic identifications of archaeobotanical material given in our source databases were controlled to ensure consistency between sources and to group taxa that cannot always be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
Taxonomic names were then matched to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023] so they could be related to modern occurrences.
The aim in both cases was to resolve taxonomic ambiguity by using a more inclusive taxonomic unit.
For example, the model discussed here as *Bolboschoenus* includes both archaeological occurrences of *B. maritimus* and *B. glaucus*, since these were not distinguished by archaeobotanists until recently [@WollstonecroftEtAl2011], as well as modern occurrences under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

@@

```{r data-n-occ} 
if (!file_exists(derived_data("flora_n_occ.Rdata"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  saveRDS(flora, derived_data("flora_n_occ.Rdata"))
}
flora <- readRDS(derived_data("flora_n_occ.Rdata"))

# Filter small samples
# ...and one very large one (why are there 50k occurrences of Avena sterilis?)
flora_all <- flora
flora <- flora |>
  filter(n_occ > min_n_occ) |>
  filter(taxon != "Avena sterilis")
```

```{r data-occ}
# Cleaning pipeline based on:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
if (!file_exists(derived_data("flora_occ.Rdata"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    }),
    # Update occ count after cleaning
    n_occ = map_int(occ, nrow)
  )

  # Refilter based on cleaned records
  flora <- filter(flora, n_occ > min_n_occ)

  saveRDS(flora, derived_data("flora_occ.Rdata"))
}

flora <- readRDS(derived_data("flora_occ.Rdata"))
```

Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other regions.

```{r data-background}
# TODO: slow - cache
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_asia_land, n = n_background),
    # occ_bg = map2(occ, bg, bind_rows)
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
```

Random Forest is a presence–absence approach to niche modelling and therefore requires not just data on where a species is present, but where it is definitely not present.
However 'absence' data is rarely available because it requires exhaustive survey.
In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of 'pseudo-absence' points.
The purpose of this sample is to inform the model about the nature of the underlying environment. 
The stochastic generation process means that some 'pseudo-absence' points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
Here we follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) uniform sample of points from across the land area of the study region.
These points are then weighted equally against the presences in the regression to produce a 'balanced Random Forest' [@ValaviEtAl2022].

@@

```{r data-climate}
# Reacquire derived_data/paleoclim if needed:
if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
  message("Redownloading Paleoclim data...")
  source(here("data", "data-paleoclim.R"))
}

climate <- read_stars(
  derived_data(
    "paleoclim", 
    paste0("paleoclim_", climate_periods$code, "_2_5m_w_asia.tif")
  ),
  along = NA
)

@@

climate <- st_redimension(climate, name = "period") |>
  split("band")
```

@@

```{r data-terrain}
# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
  message("Redownloading DEM data...")
  source(here("data", "data-dem.R"))
}

terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
                                            "srtm_15plus_w_asia_slope.tif",
                                            "srtm_15plus_w_asia_aspect.tif",
                                            "srtm_15plus_w_asia_wetness.tif")))
names(terrain) <- c("elevation", "slope", "aspect", "wetness")
```

* Terrain aspect and slope, which at high resolution perform well as proxies for solar radiation when modelling plant occurrence [@AustinVanNiel2011; @LeempoelEtAl2015]; and the topographic wetness index (TWI), which serves as a proxy for soil moisture and is particularly important in modelling arid environments [@KopeckyCizkova2010; @CamposEtAl2016; @DiVirgilioEtAl2018]. All three were derived from the SRTM15+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].

```{r data-soil}
# Reacquire derived_data/soilgrids if needed:
if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
  message("Redownloading SoilGrids data...")
  source(here("data", "data-soilgrids.R"))
}

@@

```{r data-environment}
# Prepare environmental data cube for prediction (not training)
# Duplicate time-invariant predictors to create dummy period dimension
terrain_cube <- list(terrain) |>
  rep(nrow(climate_periods)) |>
  set_names(climate_periods$code) |>
  c(list(along = "period")) |>
  do.call(what = c)

soil_cube <- list(soil) |>
  rep(nrow(climate_periods)) |>
  set_names(climate_periods$code) |>
  c(list(along = "period")) |>
  do.call(what = c)

>>>>

<!--
TODO before publication:
- [ ] Revise discussion of archaeo verification
- [ ] Outstanding TODOs
- [ ] Add missing references
- [x] Appendix with all hindcast predictions (HTML?)
-   [] TIFs?
- [ ] Clean up package code
- [ ] Zenodo release
- [ ] Final proofread
-->

@@

w_eurasia <- st_bbox(c(xmin = -30, xmax = 90, ymin = 0, ymax = 60), crs = 4326)

@@

```{r define-parameters}
# Minimum number archaeological occurrences for modelling
min_n_archaeo <- 3

# Minimum number of present occurrences for modelling
min_n_occ <- 40

@@

Biogeographic research into the centres of origin and/or domestication of crops has also long informed broader understanding of the process of agricultural origins.
In *Man Makes Himself*, @Childe1936 correctly located the centre of origin of European agriculture in the 'Fertile Crescent' of West Asia (unlike for example @Pumpelly1908 before him).
This was not based on the region's prehistoric archaeological record, which at the time had only been cursorily explored.
Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].
In subsequent decades, the search for more precise origin zones of specific domestic plants relied on the assumption that "the locus of domestication of a wild plant would presumably be within its area of original distribution in the wild state" [@Butzer1971; paraphrasing @Helbaek1959] – and that this "natural habitat" has not changed significantly over the last 12,000 years [@Butzer1971].

@@

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguished from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@@

ne_lakes <- read_sf(raw_data("ne"), "ne_10m_lakes") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_eurasia, 10)) |>
  st_union()

ne_ocean <- read_sf(raw_data("ne"), "ne_10m_ocean") |>
  st_geometry() |>
  st_crop(buffer_bbox(w_eurasia, 10)) |>
  st_union()

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_eurasia, 10)) |>
  st_difference(ne_ocean) |>
  # st_difference(ne_lakes) |>
  st_union()

s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)
w_asia_land <- st_intersection(st_as_sfc(w_asia), ne_land)
w_eurasia_land <- st_intersection(st_as_sfc(w_eurasia), ne_land)

@@

```{r data-flora}
# Start with flora present at archaeological sites
if (!file_exists(derived_data("flora.Rdata"))) {
  flora <- read_tsv(
    raw_data("swasia_neolithic_flora.tsv"),
    col_types = cols(
      source = col_character(),
      source_id = col_character(),
      source_site_name = col_character(),
      site_name = col_character(),
      latitude = col_double(),
      longitude = col_double(),
      phase = col_character(),
      phase_description = col_character(),
      phase_code = col_character(),
      age_start = col_integer(),
      age_end = col_integer(),
      taxon_source = col_character(),
      n = col_double(),
      prop = col_double(),
      reference = col_character(),
      taxon_detail = col_character(),
      taxon = col_character(),
      genus = col_character(),
      family = col_character(),
      category = col_character(),
      founder_crop = col_character(),
      edibility = col_character(),
      grass_type = col_character(),
      legume_type = col_character()
    )
  ) |>
    rename(taxon_group = "taxon", taxon = "taxon_detail") |>
    # Assign to archaeological periods
    mutate(
      age_mid = age_end + ((age_start - age_end) / 2),
      period = cut(
        -age_mid, 
        breaks = -c(archaeo_periods$start_bp, 0), 
        labels = archaeo_periods$period,
        ordered_result = TRUE
      ),
      climate_period = cut(
        -age_mid, 
        breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
        labels = climate_periods[climate_periods$code != "cur",]$code,
        ordered_result = TRUE
      )
    ) |>
    # Filter to Epipal/Neo
    filter(period %in% epipal | period %in% ppn) |>
    # Group by taxon
    drop_na(taxon) |>
    nest(.by = c(taxon), .key = "archaeo") |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")      )
    ) |>
    # Normalise taxon names and match to the GBIF Backbone Taxonomy
    mutate(
      taxon_original = taxon,
      # Spot corrections
      taxon = case_match(
        taxon,
        # c("Bolboschoenus glaucus", "Bolboschoenus maritimus", "Scirpus maritimus") ~ "Bolboschoenus",
        "Vicia narbonense" ~ "Vicia narbonensis", # N=3
        "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
        "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
        "Lens orientalis" ~ "Vicia orientalis", # N=2
        "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
        "Stipa capensis" ~ "Stipa dregeana", # N=2
        .default = taxon
      ),
      # Recode domesticates as their wild progenitors
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Lens culinaris" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      ),
      taxon_gbif = name_backbone_checklist(taxon, kingdom = "Plantae"),
      .before = archaeo
    ) |>
    mutate(
      # taxon = coalesce(taxon_gbif$species, taxon_gbif$genus),
      taxon = taxon_gbif$species, # not interested in genus-only matches
      gbif_key = coalesce(taxon_gbif$acceptedUsageKey, taxon_gbif$usageKey),
      .before = archaeo
    ) |>
    # Drop taxons not matched
    drop_na(taxon) |>
    # Regroup by canonical taxon
    select(-taxon_gbif) |>
    summarise(
      gbif_key = first(gbif_key),
      taxon_aliases = list(taxon_original[taxon_original != taxon]),
      # TODO: this creates duplicate rows
      archaeo = list(bind_rows(archaeo)),
      .by = taxon
    ) |>
    mutate(
      # Discard duplicates arising from merging of taxonomic synonyms
      archaeo = map(archaeo, function(x) {
        if (ncol(x) > 0) distinct(x, phase_code, .keep_all = TRUE)
        else data.frame()
      }),
      n_archaeo = map_int(archaeo, nrow)
    ) |>
    filter(n_archaeo >= min_n_archaeo)

@@

```{r fig-region}
#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
# TODO: Update to show Eurasia (with zoom-in for sites?)
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, fill = NA) +
  layer_spatial(archaeo_sites, size = 0.5) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Based on current understandings, we assume that plant-based subsistence during this period was broad, geographically- and temporally-varied, and reflects a gradual, geographically decentred, and nonlinear tranistion to greater reliance on cultivars (i.e. agriculture, see @sec-bg).
Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identified species observed at at least `r english(min_n_archaeo)` Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@LucasFuller2018; @FullerEtAl2018; based on @ColledgeEtAl2004; @ShennanConolly2007].
We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

The taxonomic identifications of archaeobotanical material given in our source databases were previously controlled to ensure consistency between sources and to remove taxa that cannot be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
Taxonomic names were then matched to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023] so they could be related to modern occurrences.
Archaeologically-attested domestic species meeting our inclusion criteria were substituted for their wild progenitors (where different) when gathering occurrence data, since the domestic forms are now widespread and presumably uninformative of the species original niche.

@@

```{r data-occ}
# Download occurrence data from GBIF, generating a citable derived dataset
if (length(dir_ls(derived_data("gbif"))) == 0) {
  # Requires GBIF credentials as environment variables, see:
  # https://docs.ropensci.org/rgbif/articles/gbif_credentials.html
  message("Reacquiring GBIF occurrences...")
  gbif_download <- occ_download(
    pred_in("taxonKey", flora$gbif_key),
    pred_within(wk::as_wkt(w_eurasia)),
    pred_default(), # Filter common data quality issues, see ?rgbif::pred_default() for details
    # pred_lt("coordinateUncertaintyInMeters", 1000), # to strict for W. Asia!
    format = "SIMPLE_CSV"
  ) |>
    occ_download_wait() # wait for dataset to be prepared

  # Cache data
  occ_download_get(gbif_download$key, path = derived_data("gbif"))

  # Cache metadata
  gbif_download |>
    unclass() |>
    list_flatten() |>
    discard_at(\(x) str_starts(x, "request")) |>
    as_tibble() |>
    write_csv(derived_data(
      "gbif",
      paste0(gbif_download$key, "_metadata.csv")
    ))
}

flora_occ <- dir_ls(derived_data("gbif"), glob = "*.zip")[1] |> # assume only one dataset present
  as.download() |>
  occ_download_import() |>
  gbif_standardise_names() |>
  select(taxon_key, gbif_id, latitude, longitude) |>
  group_by(taxon_key) |>
  distinct(latitude, longitude, .keep_all = TRUE)

flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1])

# Join to archaeo data
flora <- flora_occ |>
  nest(.key = "occ") |>
  ungroup() |>
  right_join(flora, by = c("taxon_key" = "gbif_key")) |>
  mutate(n_occ = map_int(occ, vctrs::vec_size))

# Filter small samples
flora_all <- flora
flora <- flora |>
  filter(n_occ >= min_n_occ)
```

<!-- TODO: cite dataset -->
Georeferenced occurrence data for West Eurasia between 0 and 60° of latitude was obtained from the Global Biodiversity Information Facility (GBIF) using via and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
The GBIF dataset [@GBIFDerivedData] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we did not attempt to model `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
Multiple records of the same taxon at the same coordinate were discarded because they do not impart information to the model.
The resulting cleaned dataset used to train our niche models comprises `r flora_occ_meta$totalRecords` from `r flora_occ_meta$numberDatasets` constituent datasets.

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
# TODO: Update, Eurasia
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia.

```{r data-background}
# TODO: slow - cache
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_eurasia_land, n = n_background),
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
```

Random Forest is a presence–absence approach to niche modelling and therefore requires not just data on where a species is present, but where it is definitely not present.
However 'absence' data is rarely available because it requires exhaustive survey.
In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of points.^[We use a background sample that is representative of the entire study area, as opposed to attempting to construct a 'pseudo-absence' sample. See @SilleroEtAl2021 for the distinction.]
The purpose of this sample is to inform the model about the nature of the underlying environment. 
The stochastic generation process means that some of these points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
Here we follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) uniform sample of points from across the land area of the study region.
These points are then weighted equally against the presences in the regression to produce a 'balanced Random Forest' [@ValaviEtAl2022].

@@

```{r data-climate}
# Reacquire paleoclim data if needed:
if (!file_exists(derived_data("climate.nc"))) {
  if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
    message("Redownloading Paleoclim data...")
    source(here("analysis", "data", "data-paleoclim.R"))
  }

  # Store as multidim file so we can read back in as a virtual raster
  message("Rebuilding palaeoclimate array...")
  climate <- read_stars(
    derived_data(
      "paleoclim", 
      paste0("paleoclim_", climate_periods$code, "_2_5m_w_eurasia.tif")
    ),
    along = NA
  )

@@

  climate <- st_redimension(climate, name = "period") |>
    split("band")

  write_mdim(climate, derived_data("climate.nc"))
}

climate <- read_stars(derived_data("climate.nc"), proxy = TRUE)
# Recover dimension values lost in NC conversion
climate <- st_set_dimensions(climate, "period", values = climate_periods$code)
```

@@

```{r data-terrain}
# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
if (!file_exists(derived_data("dem", "srtm_15plus_w_eurasia.tif"))) {
  message("Redownloading DEM data...")
  source(here("analysis", "data", "data-dem.R"))
}

terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_eurasia.tif", 
                                            "srtm_15plus_w_eurasia_slope.tif",
                                            "srtm_15plus_w_eurasia_aspect.tif",
                                            "srtm_15plus_w_eurasia_wetness.tif")))
names(terrain) <- c("elevation", "slope", "aspect", "wetness")
```

* Terrain aspect and slope, which at high resolution perform well as proxies for solar radiation when modelling plant occurrence [@AustinVanNiel2011; @LeempoelEtAl2015]; and the topographic wetness index (TWI), which serves as a proxy for soil moisture and is particularly important in modelling arid environments [@KopeckyCizkova2010; @CamposEtAl2016; @DiVirgilioEtAl2018]. All three were derived from the SRTM30+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].

```{r data-soil}
# Reacquire derived_data/soilgrids if needed:
if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
  message("Redownloading SoilGrids data...")
  source(here("analysis", "data", "data-soilgrids.R"))
}

@@

```{r data-environment}
# Prepare environmental data cube for prediction (not training)
if (!file_exists(derived_data("env_w_asia.nc"))) {
  climate_w_asia <- st_crop(climate, w_asia)
  terrain_w_asia <- st_crop(terrain, w_asia)
  soil_w_asia <- st_crop(soil, st_transform(w_asia, soil))

  # Duplicate time-invariant predictors to create dummy period dimension
  terrain_cube <- list(terrain_w_asia) |>
    rep(nrow(climate_periods)) |>
    set_names(climate_periods$code) |>
    c(list(along = "period")) |>
    do.call(what = c)

  soil_cube <- list(soil_w_asia) |>
    rep(nrow(climate_periods)) |>
    set_names(climate_periods$code) |>
    c(list(along = "period")) |>
    do.call(what = c)
